Automatic Annotation of Tasks in Structured Code

### 2018-07-14 docs/ACM/paper.tex
Adding latest changes.
Adding new figures with examples.

%% For double-blind review submission
\documentclass[sigplan,10pt,review,anonymous]{acmart}
\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

%% Conference information (used by SIGPLAN proceedings format)
%% Supplied to authors by publisher for camera-ready submission
\acmConference[PL'17]{IEEE}{2018}{Planet Earh}
\acmYear{2017}
\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\fi

@@

\title[Automatic Identification and Annotation of Tasks in Structured
Programs]
{Automatic Identification and Annotation of Tasks in Structured Programs}

\author{Pedro Henrique Ramos Costa}
\authornote{with author1 note}          %% \authornote is optional;
\orcid{nnnn-nnnn-nnnn-nnnn}
\affiliation{
  \position{Researcher}
  \department{DCC}
  \institution{UFMG}
  \streetaddress{6627 Antonio Carlos Avenue}
  \city{Belo Horizonte}
  \state{Minas Gerais}
  \postcode{31.270-213}
  \country{Brazil}
}
\email{pedroramos@dcc.ufmg.br}

\author{Gleison Souza Diniz Mendonc\c{c}a}
\authornote{with author1 note}          %% \authornote is optional;
\orcid{nnnn-nnnn-nnnn-nnnn}
\affiliation{
  \position{Researcher}
  \department{DCC}
  \institution{UFMG}
  \streetaddress{6627 Antonio Carlos Avenue}
  \city{Belo Horizonte}
  \state{Minas Gerais}
  \postcode{31.270-213}
  \country{Brazil}
}
\email{gleison.mendonca@dcc.ufmg.br}

\author{Divino C\'{e}sar}
\authornote{with author1 note}
\orcid{nnnn-nnnn-nnnn-nnnn}
\affiliation{
  \position{Researcher}
  \department{IC}
  \institution{UNICAMP}
  \streetaddress{Cidade Universit\'{a}ria}
  \city{Campinas}
  \state{S\~{a}o Paulo}
  \postcode{13083-970}
  \country{Brazil}
}
\email{divcesar@gmail.com}

\author{Guido Ara\'{u}jo}
\authornote{with author1 note}
\orcid{nnnn-nnnn-nnnn-nnnn}
\affiliation{
  \position{Professor}
  \department{IC}
  \institution{UNICAMP}
  \streetaddress{Cidade Universit\'{a}ria}
  \city{Campinas}
  \state{S\~{a}o Paulo}
  \postcode{13083-970}
  \country{Brazil}
}
\email{guido@ic.unicamp.br}

\author{Fernando Magno Quint\~{a}o Pereira}
\authornote{with author1 note}
\orcid{nnnn-nnnn-nnnn-nnnn}
\affiliation{
  \position{Professor}
  \department{DCC}
  \institution{UFMG}
  \streetaddress{6627 Antonio Carlos Avenue}
  \city{Belo Horizonte}
  \state{Minas Gerais}
  \postcode{31.270-213}
  \country{Brazil}
}
\email{fernando@dcc.ufmg.br}          %% \email is recommended

@@

Annotation systems have risen to a place of prominence as a simple and
effective means to write parallel programs.
Examples of such systems include OpenMP~\cite{JaegerCP15},
OpenACC~\cite{OpenACC20}, OpenHMPP~\cite{Andion14}, OpenMPC~\cite{Lee10},
OpenSs~\cite{MeenderinckJ11}, Cilk++~\cite{Leiserson09} and
OmpSs~\cite{bueno:icpp:2011, duran:ppl:2011}.
Annotations work as a meta-language: they let developers grant parallel
semantics to syntax originally written to execute sequentially.
%\guido{What do you mean by modern accelerators? Not clear.}
Combined with modern hardware accelerators such as GPUs and FPGAs, they have led
to substantial performance gains~\cite{Bertolli14,Mendonca17,Poesia17}.
%\guido{Suppport what?}
Nevertheless, although convenient, the use of annotations is not straightforward, 
and still lacks supporting tools that help programmers to check if annotations
are correct and/or effective.
%replaced:
%Yet, even if convenient,

%Annotations such as OpenMP or OpenACC still require developers to
%worry about typical hassles of the parallel world, such as race conditions
%and deadlocks.
%Moreover, because they are used in tandem with imperative programming languages,
%such troubles are only made worse by pointer aliasing.
There exist tools that insert automatic annotations in
programs~\cite{Mendonca16,Pingali11}.
All these technologies explore data-parallelism -- the possibility of running
the same computation independently on different data.
Another form of parallelism, based on tasks, remains still uncharted land in what
concerns automatic annotation.
Such fact is unfortunate, since much of the power of current annotation
systems lies in their ability to create tasks~\cite{Ayguade09}.
%\guido{Why the ability to create tasks makes such programs similar to irregular code?
%Which features of irregular code are resposible for such similarities? Not clear to the 
%reader}
As we illustrate in Section~\ref{sub:adv}, task parallelism -- the power to run
different routines simultaneously on independent data -- brings annotation systems
closer to irregular programs such as those that process graphs and
worklists~\cite{Pingali11}.
The purpose of this work is to address this omission.

@@

The main advantage of using OpenMP annotations to split a program into parallel
tasks is
the ability  of leaving the job of  handling dependences to a runtime.
The OpenMP runtime maintains a task dependence graph which dynamically exposes more parallelism 
opportunities than those resulting from a conservative  compile-time  analysis.
In particular, the runtime lets us circumvent the shortcomings that pointer aliasing
impose on the automatic parallelization of code.
Aliasing -- the possibility of two pointers dereferencing the same
memory region -- either prevents parallelism altogether, or forces compilers to
resort to complex runtime checks to ensure its
correctness~\cite{Alves15,Ravishankar14}.
The OpenMP runtime shields us from this problem, because
it already checks for dependencies among tasks, and dispatches them in a
correct order~\cite{LaGrone11}.

@@

We introduce Challenge~\ref{ch:cutoff} by quoting Duran {\em et al.}:
``{\em In task parallel languages, an important factor for achieving a good
performance is the use of a cut-off technique to reduce the number of tasks
created}"~\cite{Duran08b}.
This observation is particularly true in the context of recursive, fine-grained
tasks, as we analyze in Section~\ref{sub:ir}.
Figure~\ref{fig:ex_cutoff} provides an example.
To place a limit on the number of tasks simultaneously in flight, we associate
the invocation of recursive functions annotated with task pragmas with a
counter -- \textsf{taskminer\_depth\_cutoff} in Figure~\ref{fig:ex_cutoff}.
The guard in line 7 ensures that we never exceed \textsf{DEPTH\_CUTOFF}, a
predetermined threshold.
This example, together with Figure~\ref{fig:ex_Regions}, lets us emphasize that
the code generation algorithms presented in this paper are
parameterized by constants such as \textsf{DEPTH\_CUTOFF}, or
\textsf{WORK\_CUTOFF} in Figure~\ref{fig:ex_Regions}. Although these
have default estimates, we provide them as parameters to be set at will.

@@

\paragraph{Bounding Recursive Tasks}
During the evaluation of the \Taskminer, we observed that recursive
programs experienced performance regressions due to the excessive creation
of tasks.
To avoid this kind of slowdown, we currently give users the possibility to
bound the number of threads ever in flight, via a command line option, e.g.,
\textsf{./Taskminer -r 12} will limit the number of tasks to 12.
We implement this feature directly at the source code level, as part of the
final annotation of code.
Our solution is simple, yet, as the reader shall perceive in
Section~\ref{sec:eval}, it brings non-negligible benefits onto recursive
benchmarks.
Task bounding is implemented via a global variable, statically linked in the
program.
This variable is \textsf{taskminer\_depth\_cutoff} in Figure~\ref{fig:ex_cutoff}.
We insert code to increment it at the beginning of recursive functions that
are invoked within task regions, and we insert code to decrement it at each
return point of said functions.

@@

Figure~\ref{fig:TM_Performance} compares the runtime of programs produced by
\Taskminer{} against either the original program
or versions of said programs annotated manually.
In this experiment we use the benchmarks \textsf{BSC-Bots}~\cite{Duran09} (\textsf{fft} to \textsf{jacobi}
in Fig.~\ref{fig:TM_Performance}) and \textsf{Swan}~\cite{Moreira17}
(\textsf{dfs} to \textsf{private}).
Both these benchmarks come with sequential and parallel (manually
annotated) versions.
Baseline and annotated sources are compiled with gcc-6.

>>>>

%% For double-blind review submission
\documentclass[sigplan,10pt,review]{acmart}
\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

%% Conference information (used by SIGPLAN proceedings format)
%% Supplied to authors by publisher for camera-ready submission
\acmConference[PACT]{ACM}{2018}{Limassol, Cyprus}
\acmYear{2017}
\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\fi

@@

\title[Automatic Annotation of Tasks in Structured
Code]
{Automatic Annotation of Tasks in Structured Code}

\author{Pedro Ramos, Gleison Souza}
\orcid{nnnn-nnnn-nnnn-nnnn}
\affiliation{
%  \position{Researcher}
%  \department{DCC}
  \institution{UFMG, Brazil}
%  \streetaddress{6627 Antonio Carlos Avenue}
%  \city{Belo Horizonte}
%  \state{Minas Gerais}
%  \postcode{31.270-213}
%  \country{Brazil}
}
\email{pedroramos@dcc.ufmg.br}
\email{gleison.mendonca@dcc.ufmg.br}

%\author{Gleison Souza Diniz Mendonc\c{c}a}
%\authornote{with author1 note}          %% \authornote is optional;
%\orcid{nnnn-nnnn-nnnn-nnnn}
%\affiliation{
%  \position{Researcher}
%  \department{DCC}
%  \institution{UFMG}
%  \streetaddress{6627 Antonio Carlos Avenue}
%  \city{Belo Horizonte}
%  \state{Minas Gerais}
%  \postcode{31.270-213}
%  \country{Brazil}
%}
%\email{gleison.mendonca@dcc.ufmg.br}

\author{Divino C\'{e}sar, Guido Ara\'{u}jo}
%\authornote{with author1 note}
%\orcid{nnnn-nnnn-nnnn-nnnn}
\affiliation{
%  \position{Researcher}
%  \department{IC}
  \institution{UNICAMP, Brazil}
%  \streetaddress{Cidade Universit\'{a}ria}
%  \city{Campinas}
%  \state{S\~{a}o Paulo}
%  \postcode{13083-970}
%  \country{Brazil}
}
\email{divcesar@gmail.com}
\email{guido@ic.unicamp.br}

%\author{Guido Ara\'{u}jo}
%\authornote{with author1 note}
%\orcid{nnnn-nnnn-nnnn-nnnn}
%\affiliation{
%  \position{Professor}
%  \department{IC}
%  \institution{UNICAMP}
%  \streetaddress{Cidade Universit\'{a}ria}
%  \city{Campinas}
%  \state{S\~{a}o Paulo}
%  \postcode{13083-970}
%  \country{Brazil}
%}
%\email{guido@ic.unicamp.br}

\author{Fernando Magno Quint\~{a}o Pereira}
\authornote{This work is sponsored by CNPq, CAPEs, FAPEMIG, FAPESP, LG Electronics and
the CONTINUUM ANR Project.}          %% \authornote is optional;
%\orcid{nnnn-nnnn-nnnn-nnnn}
\affiliation{
%  \position{Professor}
%  \department{DCC}
  \institution{UFMG, Brazil}
%  \streetaddress{6627 Antonio Carlos Avenue}
%  \city{Belo Horizonte}
%  \state{Minas Gerais}
%  \postcode{31.270-213}
%  \country{Brazil}
}
\email{fernando@dcc.ufmg.br}          %% \email is recommended

@@

Annotation systems have risen to a place of prominence as a simple and
effective means to write parallel programs.
Examples of such systems include OpenMP~\cite{JaegerCP15},
OpenACC~\cite{OpenACC20}, OpenHMPP~\cite{Andion14}, OpenMPC~\cite{Lee10},
OpenSs~\cite{MeenderinckJ11} and
OmpSs~\cite{bueno:icpp:2011, duran:ppl:2011}.
Annotations let developers grant parallel
semantics to syntax originally written to execute sequentially.
%\guido{What do you mean by modern accelerators? Not clear.}
Combined with hardware accelerators such as GPUs and FPGAs, they have led
to substantial performance gains~\cite{Bertolli14,Mendonca17,Poesia17}.
%\guido{Suppport what?}
Nevertheless, although convenient, the use of annotations is not straightforward.
We still lack tools that help programmers to check if annotations
are correct and/or effective.
%replaced:
%Yet, even if convenient,

%Annotations such as OpenMP or OpenACC still require developers to
%worry about typical hassles of the parallel world, such as race conditions
%and deadlocks.
%Moreover, because they are used in tandem with imperative programming languages,
%such troubles are only made worse by pointer aliasing.
There exist tools that insert automatic annotations in
programs~\cite{Mendonca16,Pingali11}.
All these technologies explore data-parallelism -- the possibility of running
the same computation independently on different data.
Task parallelism remains still uncharted land in what
concerns automatic annotation.
Such fact is unfortunate, since much of the power of current annotation
systems lies in their ability to create tasks~\cite{Ayguade09}.
%\guido{Why the ability to create tasks makes such programs similar to irregular code?
%Which features of irregular code are resposible for such similarities? Not clear to the 
%reader}
As we illustrate in Section~\ref{sub:adv}, task parallelism -- the power to run
different routines simultaneously on independent data -- brings annotation systems
closer to irregular programs such as those that process graphs and
worklists~\cite{Pingali11}.
The purpose of this work is to address this omission.

@@

OpenMP task annotations let developers leave to the runtime the job of handling data
dependences.
The OpenMP runtime maintains a dependence graph which dynamically exposes more parallelism 
opportunities than those resulting from a conservative  compile-time  analysis.
In particular, the runtime lets us circumvent the shortcomings that pointer aliasing
impose on the automatic parallelization of code.
Aliasing -- the possibility of two pointers dereferencing the same
memory region -- either prevents parallelism altogether, or forces compilers to
resort to complex runtime checks to ensure its
correctness~\cite{Alves15,Ravishankar14}.
The OpenMP runtime shields us from this problem, because
it already checks for dependencies among tasks, and dispatches them in a
correct order~\cite{LaGrone11}.

@@

We introduce Challenge~\ref{ch:cutoff} by quoting Duran {\em et al.}:
``{\em In task parallel languages, an important factor for achieving a good
performance is the use of a cut-off technique to reduce the number of tasks
created}"~\cite{Duran08b}.
This observation is particularly true in the context of recursive, fine-grained
tasks, as we analyze in Section~\ref{sub:ir}.
Figure~\ref{fig:ex_cutoff} provides an example.
To place a limit on the number of tasks simultaneously in flight, we associate
the invocation of recursive functions annotated with task pragmas with a
counter -- \textsf{cutoff} in Figure~\ref{fig:ex_cutoff}.
The guard in line 7 ensures that we never exceed \textsf{DEPTH\_CUTOFF}, a
predetermined threshold.
This example, together with Figure~\ref{fig:ex_Regions}, lets us emphasize that
the code generation algorithms presented in this paper are
parameterized by constants such as \textsf{DEPTH\_CUTOFF}, or
\textsf{WORK\_CUTOFF} in Figure~\ref{fig:ex_Regions}. Although these
have default estimates, we provide them as parameters to be set at will.

@@

\paragraph{Bounding Recursive Tasks}
During the evaluation of the \Taskminer, we observed that recursive
programs experienced performance regressions due to the excessive creation
of tasks.
To avoid this kind of slowdown, we currently give users the possibility to
bound the number of threads ever in flight, via a command line option, e.g.,
\textsf{./Taskminer -r 12} will limit the number of tasks to 12.
We implement this feature directly at the source code level, as part of the
final annotation of code.
Our solution is simple, yet, as the reader shall perceive in
Section~\ref{sec:eval}, it brings non-negligible benefits onto recursive
benchmarks.
Task bounding is implemented via a global variable, statically linked in the
program.
This variable is \textsf{cutoff} in Figure~\ref{fig:ex_cutoff}.
We insert code to increment it at the beginning of recursive functions that
are invoked within task regions, and we insert code to decrement it at each
return point of said functions.

@@

Figure~\ref{fig:TM_Performance} compares the runtime of programs produced by
\Taskminer{} against either the original program
or versions of said programs annotated manually.
In this experiment we use the benchmarks \textsf{BSC-Bots}~\cite{Duran09} (\textsf{fft} to \textsf{jacobi}
in Fig.~\ref{fig:TM_Performance}) and \textsf{Swan}~\cite{Moreira17}
(\textsf{dfs} to \textsf{private}).
Both these benchmarks come with sequential and parallel (manually
annotated) versions.
Baseline and annotated sources are compiled with gcc-6 -O3.
